{\bf {\Large Improving estimates of ecosystem metabolism by removing effects of tidal advection in dissolved oxygen time series}}

### 2015-01-17 intro.tex
almost done with Smith revisions

Ecosystem metabolism is broadly defined as the difference between primary production and aerobic respiration and provides a basis for evaluating trophic state \citep{Kemp12,Needoba12}.  Primary producers, such as phytoplankton, establish the means of energy transfer to upper trophic levels. Productive systems are characterized by more efficient transfer of organic matter between trophic levels, whereas less productive systems are sinks of organic matter that are supported by allochthonous sources of energy input.  The balance between production and respiration is an integrated measure of metabolism that accounts for varying rates in processes that create and consume organic matter.  Although metabolic rates vary naturally in different regions \citep{Caffrey04}, human activities and infrastructure development are contributing factors that increase rates of production \citep{Diaz08}.  Inputs of limiting nutrients beyond background concentrations may decrease the resilience of an ecosystem such that higher rates of production are coupled with higher biological oxygen demand \citep{Yin04,Kemp09}.  Cultural eutrophication is frequently linked to declines in water quality through lower levels of dissolved oxygen and increased frequency of noxious algal blooms.  Reliables estimates of ecosystem metabolism are critical for measuring both background rates of production and potential impacts of human activities on ecosytem condition.

Ecosystem metabolism can be estimated using several techniques, each of which is appropriate under different conditions or assumptions.   Bottle-based techniques rely on rate measurements from discrete water quality samples, whereas open-water techniques infer metabolic rates using \textit{in situ} measurements from continuous monitoring data.  Bottle-based techniques are useful for direct partitioning of metabolic contributions into discrete habitats, such as planktonic production rates during specific time periods \citep{Kemp12}.  However, such measurements may be inappropriate for evaluating whole ecosystem metabolism if significant production occurs in habitats that are not sampled, such as benthic or seagrass production.  As such, the open-water technique provides an integrative measure of metabolism by inferring process rates from \textit{in situ}, continuous monitoring data.  Originally proposed for use in streams \citep{Odum56}, the method has been used with varying success in lakes \citep{Staehr10,Coloso11,Batt12} and estuaries \citep{Caffrey04,Hopkinson05,Caffrey13}.  The ability of the open-water method to accurately estimate metabolism depends on whether the assumptions for its use are met, which are often only implicity verified in practiced.

The open-water method uses the diel fluctuation of dissolved oxygen to infer rates of ecosytem metabolism, after correcting for losses or gains through air-water exchange \citep{Kemp12}.  Daily integrated measurements of metabolism are based on the balance between daytime estimates of gross production and nighttime estimates of respiration extrapolated to a 24 hour period.  The fundamental assumption of the open-water method is that measurements come from a water mass that has the same recent history \citep{Needoba12}.  Estimates of metabolism from a single location may be inaccurate if substantial variation in water column mixing occurs throughout the period of observation.  As such, the original technique designed for use in streams requires the comparison of data from an upstream and downstream station \citep{Odum56}.  Application of the method to systems without continuous flow, such as lakes or estuaries, have often assumed that a single sampling station provides sufficient data for estimating metabolism \cite{Staehr10}.  While single stations may be valid under specific conditions, numerous studies have shown that the open-water method may be inappropriate given the effects of physical mixing \citep{Ziegler98,Caffrey03,Coloso11,Batt12,Nidzieko14}.

The open-water method has recently been applied to coastal and oceanic ecosystems with mixed success.  An exhaustive analysis by \citet{Caffrey03} applied the method to estimate metabolim at 28 continuous monitoring stations at 14 US estuaries.  Data from two of the reserves were used to evaluate the assumption of homogeneity of water masses measured by each sensor.  Although significant differences were not observed for metabolism estimates between adjacent stations, the analysis was based on a comparison of means using conventional significance tests rather than a systematic comparison of time series.  Moreover, a portion of metabolism estimates from all stations were negative for production during the day and positive for respiration during the night.  These values were opposite in sign than expected since production increases oxygen during the day (i.e., positive effect on metabolism) and respiration consumes oxygen at night (i.e., negative effect on metabolism).  These `anamolous' values were attributed to violations in the assumption of water-column heterogeneity.  Specifically, tidal variation could have caused sampling of different water masses by individual water quality sondes as water moved inland or seaward with changing tide.

The effects of tidal advection on estimates of ecosystem metabolism have been a point of concern in numerous studies \citep{Ziegler98,Caffrey03,Collins13,Howarth14}, although systematic estimates of its effects and methods for accounting for physical variation in \ac{DO} measurements have been minimal.  An exception is presented by \citet{Nidzieko14} through quantitative assessment of the effects of fortnightly tidal modulations on metabolism estimates.  Using a control volume approach to measure fluxes into and out of a shallow tidal creek, significant biases in metabolism estimates were observed.  Net heterotrophy was observed during spring tides, whereas metabolism was balanced during neap tides.  The timing of irradiance relative to the tidal cycle was a primary factor contributing to heterotrophy during summer months such that maximum tides occurred during the night, increasing total area for respiration.  The results of the analysis, although specific to the study location, suggest that the effects of tidal advection on \ac{DO} measurements are of primary concern when selecting locations and length of time for sonde deployment in estuaries.  In many cases, the relative magnitude of these effects may be a significant source of bias without quantitative evaluation to determine the roles of biological and physical signals in \ac{DO} measurements. Analytical techniques to evaluate and correct for tidal advection could improve certainty in metabolism estimates and also increase the use of data from shorter deployment periods if sources of bias are quantified and removed.

This article describes use of a novel method for quantifying and removing noise in estimates of ecosystem metabolim for estuaries.  Specifically, we characterize the effects of tidal advection on \ac{DO} observations to improve estimates of open-water metabolism with multi-year time series of high frequency ($<$ one hour) water quality data.  The focus of our analysis is the use of a weighted regression method previously developed for trend analysis of pollutant concentrations in streams and rivers \citep{Hirsch10}.  A weighted regression approach is applied to create dynamic predictions of \ac{DO} as a function of time and tidal height change, which is then used to normalize, or detide, the \ac{DO} signal.  The analysis is presented in two steps.  First, we apply a simulation approach to create time series of \ac{DO} observations with known characteristics to evaluate ability of the weighted regression to predict the time series and remove the effects of tidal advection.  Second, four case studies of multi-year time series are used to further explore use of the weighted regression approach to remove potential noise in \ac{DO} signals from tidal advection.  Comparisons of observed and detided \ac{DO} values are compared, in addition to estimates of open-water metabolism before and after detiding of the \ac{DO} time series.  Overall, the analysis provides a means to improve certainty in conclusions from observed \ac{DO} for evaluating the relative roles of biological and physical processes in estuarine systems.  Applications of the weighted regression approach are expected to have wide-ranging implications for management and ecosystem monitoring by outlining strategies for obtaining water quality estimates with more accuracy.

>>>>

Ecosystem metabolism describes the balance between production and respiration processes that create and consume organic matter \citep{Kemp12,Needoba12}.  Light exposure experiments of water samples collected at discrete locations and times have traditionally been used to measure metabolic activity.  Although highly controlled and precise, bottle-based methods are labor-intensive and not scalable to describe entire ecosystem rates.  Bottle-based methods may also only reliably estimate production and respiration associated with planktonic processes, whereas significant contributions of ecosystem production can arise from other habitats, such as benthic seagrass communities.  By contrast, open-water techniques have been increasingly used to estimate whole system metabolism given the availability of long-term, continuous time series of dissolved oxygen \citep{Odum56,Davanzo96}. Daily integrated measurements of metabolism represent the balance between daytime production and nighttime respiration attributed to all ecosystem components.  Open-water estimates also provide a basis for tracking ecosystem change over time and are more practical for capturing events or evaluating trends as compared to bottle-based approaches. Although metabolic rates vary naturally at different spatial and temporal scales \citep{Ziegler98,Caffrey04,Russell07}, anthropogenic nutrient sources are often contributing factors that increase rates of production (\citealt{Nixon95}, \citetalias{NRC00}).  Reliables estimates of whole ecosystem metabolism are critical for measuring both background rates of production and potential impacts of human activities on ecosytem condition.

The ability to accurately estimate whole system metabolism using the open-water method depends on the degree to which assumptions of the theory are met \citep{Staehr10,Kemp12}.  The fundamental assumption is that the time series of \ac{DO} represents a Lagrangian specification of the flow field that describes the same water mass over time \citep{Needoba12}.  The Lagrangian specification assumes that the time series characterizes individual fluid parcels regardless of location, as in a parcel of water moving with the tide.  In reality, most \ac{DO} time series are collected at fixed locations such as a mooring or dock, which is characteristized by an Eulerian specification of the flow field.  Time series at fixed locations may characterize water masses with different metabolic histories if water particles are transported by physical advection. A Lagrangian flow field is often assumed, such that estimates of metabolism may be inaccurate if substantial variation in water column mixing occurs throughout the period of observation \citep{Kemp80,Russell07}.  Given this critical challenge, the open-water method has been used with varying success in estuaries influenced by tidal mixing \citep{Caffrey04,Russell07,Caffrey14}.  Appropriate placement of monitoring sondes, sampling frequency and duration, and reliability of data from single stations have been relevant issues in applying the open-water method to systems influenced by physical mixing \citep{Russell07,Staehr10}.  Application of the method to estuaries is a particular concern as physical mixing caused by tidal currents may confound the biological variation in \ac{DO} time series \citep{Kemp80,Caffrey03,Nidzieko14}.  Individual sampling stations near bay inlets or along major tidal axes may produce \ac{DO} time series that fail to meet the assumptions of the open-water method.

Although numerous studies have shown that application of the open-water method to lakes or estuaries may be problematic \citep{Ziegler98,Caffrey03,Coloso11,Batt12,Nidzieko14}, very few quantitative approaches have been developed to address potential bias or noise in \ac{DO} signals from physical advection.  For example, an extensive analysis by \citet{Caffrey03} applied the open-water method to estimate metabolism at 28 continuous monitoring stations at 14 US estuaries.  A significant portion of the production and respiration estimates were negative (3 - 69\% depending on site), suggesting advection of water masses was a likely factor influencing the \ac{DO} time series.  These `anomalous' values are typically omitted from the analysis \citep{Caffrey03,Collins13}, which may upwardly bias estimates of metabolism \citep{Caffrey14}.  Further, \citet{Nidzieko14} evaluated the effects of tidal advection on metabolism estimates in a mesotidal estuary.  Estimates from a single location were strongly correlated with the spring-neap cycle such that net heterotrophy was more common during spring tides, whereas metabolism was generally balanced during neap tides.  A control-volume approach was used by impounding a section of the upper estuary to understand how physical processes contribute to biological variability.  Although useful as an \textit{in situ}, site-specific approach, more accessible statistical methods specific to time series are needed given the increasing availability of continuous monitoring data. For example, \citet{Batt12} explored the use of a Kalman filter \citep{Harvey89} to remove process and observation uncertainty from \ac{DO} time series in lakes.  Similar approaches have not been developed for estuaries, particularly those that address potential effects of tidal advection.

This article describes the development and application of a method for improving estimates of ecosytem metabolism computed from \ac{DO} time series.  Specifically, the apparent effects of tidal advection on \ac{DO} observations are removed to improve the fidelity of open-water metabolism estimates derived from continuous water quality data.  We used a weighted regression approach originally developed to resolve trends in pollutant concentrations in streams and rivers \citep{Hirsch10}.  The weighted regression approach creates dynamic predictions of \ac{DO} as a function of time and tidal height change, which are then used to filter, or detide, the \ac{DO} signal.  The model is based on the recognition that daily fluctuations in \ac{DO} are caused by metabolism associated with the solar cycle, whereas other fluctuations in estuaries are likely associated with water level changes that generally exhibit pregression relative to the solar cycle.  The weighted regression model was applied, rather than methods commonly used for detiding in physical oceanography, to allow for the complex and dynamic patterns of \ac{DO} changes relative to advection.  First, we used simulated \ac{DO} time series with known characteristics to evaluate ability of the weighted regression to remove the simulated effects of a tidally-advected \ac{DO} gradient.  Second, the simulation results informed the application of the method to four case studies chosen from the \aclu{NERRS} (\acs{NERRS}, \citealt{Wenner04}).  In all examples, tidal height is used as a proxy for lateral water movements that may influence \ac{DO} observations.  In the absence of quantitative data describing lateral \ac{DO} variation (e.g., contemporaneous stations along a tidal axis), we assume tidal height is an appropriate characterization of lateral variation.  Accordingly, `tidal variation' or `changes in tidal height' are used throughout in reference to assumed lateral \ac{DO} gradients that are carried past monitoring sensors by tidal currents.

### 2015-01-17 mat_proc.Rnw
almost done with Smith revisions

Weighted regression was implemented as a moving window that allowed for estimation of \ac{DO} throughout the time series by adapting to variation through time as a function of tide. Regression models were estimated sequentially for each observation in the time series using dynamic weight vectors that change with the center of the window.  Weight vectors quantified the relevance of observations to the center of the window in respect to time, hour of the day, and tidal height.  Specifically, weights were assigned to each variable using a tri-cube weighting function \citep{Tukey77,Hirsch10}:
\begin{equation}
w= \left\{ 
  \begin{array}{l l}
    \left(1-\left(d/h\right)^3\right)^3 & \quad \textrm{if } |d| \leq h \\
    0 & \quad \textrm{if } |d| > h 
  \end{array} \right.
\end{equation}
where the weight $w$ of each observation is inversely proportional to the distance $d$ from the center of the window such that observations more similar to the point of reference are given higher importance in the regression.  Weights exceeding the maximum width of the window $h$ are equal to zero.  The tri-cube weighting function is similar to a Gaussian distribution such that weights decrease gradually from the center until the maximum window width is reached.  Regressions that use simpler windows (e.g., boxcar approach) are  more sensitive to influential observations as they enter or leave the window, whereas the tri-cube function minimizes their effect through gradual weighting of observations from the center \citep{Hirsch10}.  The final weight vector for each observation is the product of three separate weight vectors for time (day), hour, and tidal height. Windows for time and hour weight observations based on distance (time) from the center of the window.  The window for tidal height weights observations based on the difference from the center as a proportion of the total tidal height range.  For example, a half-window width of 0.5 means that observations are weighted proportionately within +/- 50\% the total range referenced to the tidal height in the center of the window. A low weight is given to an observation if any of the three weighting values were not similar to the center of the window since the final weight vector is the product of three weight vectors for each variable (see the link in the \hyperref[multi]{multimedia} section for graphical display of different weights).

>>>>

Weighted regression was implemented as a moving window that allowed for estimation of \ac{DO} throughout the time series by adapting to variation through time as a function of tide. Regression models were estimated sequentially for each observation in the time series using dynamic weight vectors that change with the center of the window.  Weight vectors quantified the relevance of observations to the center of the window in respect to time, hour of the day, and tidal height.  Specifically, weights were assigned to each variable using a tri-cube weighting function \citep{Tukey77,Hirsch10}:
\begin{equation}
w= \left\{ 
  \begin{array}{l l}
    \left(1-\left(d/h\right)^3\right)^3 & \quad \textrm{if } |d| \leq h \\
    0 & \quad \textrm{if } |d| > h 
  \end{array} \right.
\end{equation}
where the weight $w$ of each observation is inversely proportional to the distance $d$ from the center of the window such that observations more similar to the point of reference are given higher importance in the regression.  Observations that exceed the maximum width of the window $h$ are assigned a weight of zero.  The tri-cube weighting function is similar to a Gaussian distribution such that weights are decreasing gradually from the center until the maximum window width is reached.  Regressions that use simpler windows (e.g., boxcar approach) are  more sensitive to influential observations as they enter or leave the window, whereas the tri-cube function minimizes their effect through gradual weighting of observations from the center \citep{Hirsch10}.  The final weight vector for each observation is the product of three separate weight vectors for time (day), hour, and tidal height. Windows for time and hour are used to weight observations based on distance (time) from the center of the window.  The window for tidal height is used to weight observations based on the difference from the center as a proportion of the total tidal height range.  For example, a half-window width of 0.5 means that observations are weighted proportionately within +/- 50\% the total range referenced to the tidal height in the center of the window. A low weight is given to an observation if any of the three weighting values were not similar to the center of the window since the final weight vector is the product of three weight vectors for each variable (see the link in the \hyperref[multi]{multimedia} section for graphical display of different weights).

### 2015-01-17 swmp_noise.tex
almost done with Smith revisions

Time series of dissolved oxygen are increasingly used to estimate ecosystem metabolism \citep{Kemp12,Needoba12}.  Integrated measures of metabolism describe the balance between production and respiration processes that create and consume organic matter, respectively.  Although metabolic rates vary naturally at different spatial and temporal scales \citep{Ziegler98,Caffrey04,Russell07}, anthropogenic nutrient sources are often contributing factors that increase rates of production (\citealt{Nixon95}, \citetalias{NRC00}).  Inputs of limiting nutrients beyond background concentrations may decrease the resilience of an ecosystem such that higher rates of production are coupled with higher biological oxygen demand \citep{Yin04,Kemp09}.  Cultural eutrophication is frequently linked to declines in water quality through lower levels of dissolved oxygen, degradation in aquatic vegetation habitat, and increased frequency of harmful algal blooms \citep{Cloern96,Short96,Rabalais02,Diaz08}.  Reliables estimates of ecosystem metabolism are critical for measuring both background rates of production and potential impacts of human activities on ecosytem condition.

Open-water techniques have been used for decades to infer metabolic rates, particularly in the last twenty years using \textit{in situ} measurements from continuous monitoring data \citep{Odum56,Davanzo96}. Daily integrated measurements of metabolism represent the balance between daytime production and nighttime respiration.  The open-water method uses the diel fluctuation of dissolved oxygen to estimate ecosytem metabolism, after correcting for air-water gas exchange \citep{Kemp12}.  As with any method, the ability to accurately estimate whole system metabolism depends on the degree to which assumptions of the theory are met.  The fundamental assumption is that the time series of \ac{DO} represents a Lagrangian specification of the flow field that describes the same water mass over time \citep{Needoba12}.  The Lagrangian specification assumes that the time series characterizes individual fluid parcels regardless of location, as in a parcel of water moving with the tide.  In reality, most \ac{DO} time series are collected at fixed locations such as a mooring or dock, which is characteristized by an Eulerian specification of the flow field.  Time series at fixed locations may characterize water masses with different metabolic histories if water particles are transported by physical advection. A Lagrangian flow field is often assumed, such that estimates of metabolism may be inaccurate if substantial variation in water column mixing occurs throughout the period of observation \citep{Kemp80,Russell07}.  Given this critical challenge, the open-water method has been used with varying success in lakes \citep{Staehr10,Coloso11,Batt12} and estuaries \citep{Caffrey04,Russell07,Caffrey14}.  Appropriate placement of monitoring sondes, sampling frequency and duration, and reliability of data from single stations have been relevant issues in applying the open-water method to systems influenced by physical mixing \citep{Russell07,Staehr10}.  Application of the method to estuaries is a particular concern as physical mixing caused by tidal currents may confound the biological variation in \ac{DO} time series \citep{Kemp80,Caffrey03,Nidzieko14}.  Individual sampling stations near bay inlets or along major tidal axes may produce \ac{DO} time series that fail to meet the assumptions of the open-water method.

@@

The proportion of daily integrated metabolism estimates that were anomalous (negative production, positive respiration) were significantly reduced for all sites after filtering (\cref{tab:case_res}), perhaps indicating the relative effects of water movement.  Before filtering, anomalous values ranged from 0.09 (as a proportion of the total estimates, Rookery Bay) to 0.22 (Padilla Bay) for production and 0.08 (Rookery Bay) to 0.21 (Elkhorn Slough) for respiration. Anomalous values were reduced to near zero for all case studies, particularly Rookery Bay and Sapelo Island.  Metabolism estimates using filtered \ac{DO} time series had decreased mean production (-63.1 \% change from the annual mean) and respiration (-62.6 \%) for Elkhorn Slough, increased mean production (17.8 \%) and respiration (18.8 \%) for Padilla Bay, and generally unchanged mean production and respiration for Rookery Bay and Sapelo Island (\cref{tab:case_res}).  Changes in mean estimates based on filtered \ac{DO} time series suggests that the weighted regression removed variation attributed to both biological and physical processes.  The implications of these undesirable results are described below.  Decreases in the standard erorr for all metabolism estimates were observed for all cases after filtering.

>>>>

Ecosystem metabolism describes the balance between production and respiration processes that create and consume organic matter \citep{Kemp12,Needoba12}.  Light exposure experiments of water samples collected at discrete locations and times have traditionally been used to measure metabolic activity.  Although highly controlled and precise, bottle-based methods are labor-intensive and not scalable to describe entire ecosystem rates.  Bottle-based methods may also only reliably estimate production and respiration associated with planktonic processes, whereas significant contributions of ecosystem production can arise from other habitats, such as benthic seagrass communities.  By contrast, open-water techniques have been increasingly used to estimate whole system metabolism given the availability of long-term, continuous time series of dissolved oxygen \citep{Odum56,Davanzo96}. Daily integrated measurements of metabolism represent the balance between daytime production and nighttime respiration attributed to all ecosystem components.  Open-water estimates also provide a basis for tracking ecosystem change over time and are more practical for capturing events or evaluating trends as compared to bottle-based approaches. Although metabolic rates vary naturally at different spatial and temporal scales \citep{Ziegler98,Caffrey04,Russell07}, anthropogenic nutrient sources are often contributing factors that increase rates of production (\citealt{Nixon95}, \citetalias{NRC00}).  Reliables estimates of whole ecosystem metabolism are critical for measuring both background rates of production and potential impacts of human activities on ecosytem condition.

The ability to accurately estimate whole system metabolism using the open-water method depends on the degree to which assumptions of the theory are met \citep{Staehr10,Kemp12}.  The fundamental assumption is that the time series of \ac{DO} represents a Lagrangian specification of the flow field that describes the same water mass over time \citep{Needoba12}.  The Lagrangian specification assumes that the time series characterizes individual fluid parcels regardless of location, as in a parcel of water moving with the tide.  In reality, most \ac{DO} time series are collected at fixed locations such as a mooring or dock, which is characteristized by an Eulerian specification of the flow field.  Time series at fixed locations may characterize water masses with different metabolic histories if water particles are transported by physical advection. A Lagrangian flow field is often assumed, such that estimates of metabolism may be inaccurate if substantial variation in water column mixing occurs throughout the period of observation \citep{Kemp80,Russell07}.  Given this critical challenge, the open-water method has been used with varying success in estuaries influenced by tidal mixing \citep{Caffrey04,Russell07,Caffrey14}.  In contrast, the method has been more successfully applied to water quality time series in lakes, although stratification may limit estimates to specific vertical layers \citep{Staehr10,Coloso11,Batt12}. Appropriate placement of monitoring sondes, sampling frequency and duration, and reliability of data from single stations have been relevant issues in applying the open-water method to systems influenced by physical mixing \citep{Russell07,Staehr10}.  Individual sampling stations near bay inlets or along major tidal axes may produce \ac{DO} time series that fail to meet the assumptions of the open-water method.

@@

The proportion of daily integrated metabolism estimates that were anomalous (negative production, positive respiration) were significantly reduced for all sites after filtering (\cref{tab:case_res}), perhaps indicating the relative effects of water movement.  Before filtering, anomalous values ranged from 0.09 (as a proportion of the total estimates, Rookery Bay) to 0.22 (Padilla Bay) for production and 0.08 (Rookery Bay) to 0.21 (Elkhorn Slough) for respiration. Anomalous values were reduced to near zero for all case studies, particularly Rookery Bay and Sapelo Island.  Metabolism estimates using filtered \ac{DO} time series had decreased mean production (-63.1 \% change from the annual mean) and respiration (-62.6 \%) for Elkhorn Slough, increased mean production (17.8 \%) and respiration (18.8 \%) for Padilla Bay, and generally unchanged mean production and respiration for Rookery Bay and Sapelo Island (\cref{tab:case_res}).  Changes in mean estimates based on filtered \ac{DO} time series suggests that the weighted regression removed variation attributed to both biological and physical processes.  The implications of these undesirable results are described below.  Decreases in the standard erorr for all metabolism estimates were observed for all cases after filtering.  However, reduction in standard error may also occur if variation from biological processes is reduced separately or in addition to physical variation.  The extent to which this reduction in variation is related to the former is not clear, although reductions in mean values (particularly at Elkhorn Slough) suggests that biological variation may have been removed.
